Learning a functional distance between organizations with unsupervised embeddings of scientific mobility

### 2020-07-03 paper/current/main.tex
updated SemAxis section

Prestige hierarchies underpin the dynamics of scientific mobility~\autocite{deville2014career, clauset2015hierarchy}, but are these hierarchies captured in the embedding space?
We test this by exploting the geometric properties of the embedding space with the technique SemAxis~\autocite{an2018semaxis}, an approach originally designed for word embeddings but that is equally applicable here. 
SemAxis assumes the existence of an a user-defined axis, defined as the vector formed by the distance of some two vectors in the embedding space.
The choice of the axis should represent a meaningful and coherent concept.
Organization vectors are then projected onto this axis, and their position estimates their similarity to the concepts embodied by either end of the axis. 
We define prestige using the Times Ranking of World Universities, including all ranked universities in the United States;
the elite end is defined as the mean vector of the top five U.S. universities according to the rankings, whereas the non-elite axis is defined using the five bottom-ranked and geographically matched (by U.S. census region) universities.
A geographic axis is also added to aid the visualization and reinforce the importance of geography, defined as the mean vector of all universities in California and then Massachusetts.

Universities hiring and mobility networks are known to exhbit latent pretige hierarchies ~\autocite{deville2014career, clauset2015hierarchy} that should be encoded in the embedding space.
Accordingly, prestigious universities such as Stanford, Colombia, MIT, Harvard, and NYU, group towards the top of the axis (Fig.~\ref{fig:semaxis}a).
Universities at the bottom of this axis tend to be regional universities with lower national profiles.
By considering mobility between all organization types, organization embeddings offer a new means of reasoning about the prestige of organizations typically excluded from rankings such as government organizatioss (Fig.~\ref{fig:supp:semaxis_sectors}a), research institutes (Fig.~\ref{fig:supp:semaxis_sectors}b), and teaching colleges (Fig.~\ref{fig:supp:semaxis_sectors}c).

@@

The embedding space captures prestige hierarchies to such a high resoltuion that they can be used to reconstruct academic rankings. 
We derive a ranking using the prestige axis that we call \textit{SemRank}, and compare them to traditional university rankings.
SemRank is defined from the position of organizations along the prestige axis, with the top ranked being the furthest towards the \textit{Elite} end of the axis.
We find that SemRank derived is highly correlated with the Times ranking ($\rho = 0.73$, Fig.~\ref{fig:semaxis}b).
This correlation is robust with the number of organizations used to define the axes (Fig. SI~\ref{fig:supp:semaxis_compare}), such that even using only the single top-ranked and bottom-ranked universities is strongly correlated with the Times Ranking ($\rho = 0.46$, Fig.~\ref{fig:supp:semaxis_compare}a).
For reference, SemRank achieves correlation with the Times Ranking comparable to that between the Times ranking and the Leiden Ranking~\autocite{waltman2012leidenrankings}, a common and bibliometrically-informed university ranking ($\rho = 0.87$, Fig.~\ref{fig:supp:times_vs_leiden}).
This suggests that the organization embedding captures the latent prestige hierarchy inherent to academic mobility with a high degree of resolution and reliability.

>>>>

Prestige hierarchies underpin the dynamics of scientific mobility~\autocite{deville2014career, clauset2015hierarchy}, but are these hierarchies captured in the embedding space?
We test this by exploting the geometric properties of the embedding space with the technique SemAxis~\autocite{an2018semaxis}, an approach originally designed for characterize the semantics of words in language embeddings, but is equally applicable here. 
In the context of organization vectors, this approach leverages the unique properties of the embedding space in order to quantify the nearness of an organization to the ends of some meaningful axis. 
The axis is defined by the user and should represent some meaningful  and coherent concept. 
Organization vectors are then projected onto this axis, and their position estimates their similarity to the concepts embodied by either end of the axis. 
Here, we use SemAxis to operationalize the abstract notion of academic prestige, defining each end of the axis using known high- and low-ranked universities. 
We define prestige using the Times Ranking of World Universities, including all ranked universities in the United States;
the elite end is defined as the mean vector of the top five U.S. universities according to the rankings, whereas the non-elite axis is defined using the five bottom-ranked and geographically matched (by U.S. census region) universities.
A geographic axis is also added to aid the visualization and reinforce the importance of geography, defined as the mean vector of all universities in California and then Massachusetts.

Universities hiring and mobility networks are known to exhbit latent pretige hierarchies ~\autocite{deville2014career, clauset2015hierarchy} that should be encoded in the embedding space.
Accordingly, prestigious universities such as Stanford, Colombia, MIT, Harvard, and NYU are positioned towards the top of the axis (Fig.~\ref{fig:semaxis}a).
Universities at the bottom of this axis tend to be regional universities with lower national profiles.
By considering mobility between all organization types, SemAxis applied to embeddings offers a new means of reasoning about the prestige of organizations typically excluded from rankings such as government organizatioss (Fig.~\ref{fig:supp:semaxis_sectors}a), research institutes (Fig.~\ref{fig:supp:semaxis_sectors}b), and teaching colleges (Fig.~\ref{fig:supp:semaxis_sectors}c).

@@

The embedding space captures prestige hierarchies to such a high resolution that they can be used to reconstruct academic rankings. 
We derive a ranking using the prestige axis that we call \textit{SemRank}, and compare them to traditional university rankings.
SemRank is defined from the position of organizations along the prestige axis, with the top ranked being the furthest towards the \textit{Elite} end of the axis.
We find that SemRank derived is highly correlated with the Times ranking ($\rho = 0.73$, Fig.~\ref{fig:semaxis}b).
This correlation is robust with the number of organizations used to define the axes (Fig. SI~\ref{fig:supp:semaxis_compare}), such that even using only the single top-ranked and bottom-ranked universities is strongly correlated with the Times Ranking ($\rho = 0.46$, Fig.~\ref{fig:supp:semaxis_compare}a).
For reference, SemRank achieves correlation with the Times Ranking comparable to that between the Times ranking and the Leiden Ranking~\autocite{waltman2012leidenrankings}, a common and bibliometrically-informed university ranking ($\rho = 0.87$, Fig.~\ref{fig:supp:times_vs_leiden}).
This suggests that the organization embedding captures the latent prestige hierarchy inherent to academic mobility with a high degree of resolution and reliability.

While we use SemAxis to operationalize notions of prestige and geography, the approach can be easily generalized to other concepts so long as they can be meaningfully represented as an axis in the vector space. 
For example, axes could similarly be defined between teaching- and research-focused universities, between companies and government agencies, or between organizations in the U.S. and in China.
